Cryptobenthic reef assemblages are more distinct across a 90 m depth gradient than 2,500 km of shallow marine habitat in the Hawaiian Archipelago

### 2025-03-12 manuscript.Rmd
little abstract tweaks

Investigations into the environmental and anthropogenic factors that shape coral reef communities have been limited in both their depth range and taxonomic scope. Despite high biodiversity and the known importance of mesophotic habitats, due to logistical challenges, previous studies have focused largely on fishes and corals and the shallowest ~30 m of their depth ranges. Here, we examine the variability patterns of brachyuran crabs recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth. We examine the effects of a suite of environmental, ecological, and anthropogenic factors on shallow (8--17 m) crab assemblages and compare them with those sampled along a shallow-to-mesophotic depth gradient (12--90 m) on O&#x02BB;ahu. Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, island slope, the potential for larval immigration, and depth. However, we found more variation along the depth gradient on O&#x02BB;ahu than among the shallow assemblages across the entire archipelago. This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire span of the Hawaiian Islands.

>>>>

Investigations into the environmental and anthropogenic factors that shape coral reef communities have historically been limited in both depth range and taxonomic scope. Despite high biodiversity and the recognized importance of mesophotic habitats and due to logistical challenges, previous studies have focused largely on fishes and corals and the shallowest ~30 m of their depth ranges. Here, we examine the variability patterns of brachyuran crabs recovered from Autonomous Reef Monitoring Structures across the ~2,500 km Hawaiian Archipelago and ~90 m of depth. We examine the effects of a suite of environmental, ecological, and anthropogenic factors on shallow (8--17 m) crab assemblages and compare them with those sampled along a shallow-to-mesophotic depth gradient (12--90 m) on O&#x02BB;ahu. Shallow assemblages were significantly different between the Main and Northwest Hawaiian Islands and significantly associated with human impacts, chlorophyll-A, sea-surface temperature, island slope, the potential for larval immigration, and depth. However, we found more variation along the depth gradient on O&#x02BB;ahu than among the shallow assemblages across the entire archipelago. This finding suggests that 90 m of depth is a stronger determinant of brachyuran assemblage structure than the latitudinal, environmental, or anthropogenic gradients across the entire span of the Hawaiian Islands.
